Expression QTL hotspot dissection with mediation analysis and pleiotropy testing

### 2019-10-02 analysis/paper/Rmd/abstract.Rmd
updated abstr & methods

We present our analysis of five expression trait hotspots in pancreatic islets from 378
Diversity Outbred mice.

>>>>

We present our analysis of five expression trait hotspots in pancreatic islets from 378
Diversity Outbred mice. @keller2018genetic defined the hotspots in their initial report. We now apply our newly developed pleiotropy test and mediation analysis to understand biological relationships among traits that map to the hotspots.

Four of five hotspots show patterns consistent with a single local trait causing the hotspot associations. For these hotspots, mediation analyses revealed that the local trait mediates most of the associated nonlocal traits, while pleiotropy test results were consistent with presence of a single QTL at each hotspot. However, the fifth hotspot demonstrated distinct patterns in both mediation analyses and pleiotropy testing.

### 2019-10-02 analysis/paper/Rmd/methods.Rmd
updated abstr & methods

In analyzing the data from @keller2018genetic, we first established criteria for identifying local and 
nonlocal transcripts at each of the five hotspots. We defined a "local trait" to be one whose corresponding gene lay within the hotspot. We focused on those local 
traits that strongly associated with QTL within the hotspot. To limit the scope of our analyses, 
we only studied those expression traits with univariate LOD greater than 15.

We defined a "nonlocal trait" to be one with univariate evidence of association (LOD at least 7.18, the genome-wide threshold)
with a QTL in the hotspot. Additionally, we required that the nonlocal trait's gene not lie on the hotspot's chromosome. We ignored transcripts for genes on the hotspot chromosome 
but outside the hotspot interval.

For each hotspot, we performed both pairwise pleiotropy tests and linear regression-based 
mediation analyses. For every pair of traits at a hotspot, we calculated a pleiotropy test 
statistic[@boehm2019testing]. Due to the computational burden, we didn't determine bootstrap p-values for these 
tests. For every local trait - nonlocal trait pair, we also performed a linear 
regression-based mediation analysis [@chick2016defining,@keller2018genetic]. We describe both procedures below.

@@

For $n$ mice, $\Sigma$ is a $2n$ by $2n$ covariance matrix, while $V_g$ and $V_e$ are $2$ by $2$ covariance matrices. Estimates $\hat V_g$ and $\hat V_e$ are obtained via an expectation-maximization algorithm [@zhou2014efficient;@gemma2].

From the two-dimensional scan results, we calculated pleiotropy likelihood ratio test (LOD) statistics for each
pair of traits [@jiang1995multiple;@boehm2019testing;@boehm2019thesis]. Because of the required
computing time for bootstrap analyses, we didn't obtain p-values for the pleiotropy tests. Instead,
we worked directly with the test statistic values.

@@

In all of our mediation analyses, the nonlocal traits are the targets (Y), while the local
transcripts serve as candidate mediators (M) of the QTL-target relationship. 
We only considered as potential mediators those local traits with strong univariate LODs.
In Equations~\ref{eq:model1} through \ref{eq:model4}, $X$ is a matrix of founder allele dosages,
$B$ is a matrix
of founder allele effects, $W$ is a matrix of covariates, $C$ is a matrix of covariate
effects, $M$ is a vector of values for a single candidate mediator (a local expression trait, in our case), $\beta$ is the effect
of the mediator on the complex trait (after adjusting for other covariates in the model), 1 is a vector with all entries set to 1, $\alpha$
is a parameter that corresponds to the mean trait value (after adjusting for other covariates in the model), and $E$ is a vector of normally distributed random errors.

@@

We created heatmaps to examine patterns among pleiotropy test results in each hotspot. Our analysis used both 
local and nonlocal traits.
We represented the pleiotropy test statistics as a symmetric matrix, where each row was one expression
trait and each column was an expression trait. 
Excepting the cells on the diagonal, each cell, then, contained the pleiotropy test statistic
from the analysis involving the row's expression trait and the column's expression trait.
To create a heatmap, we first performed two hierarchical clusterings of rows. 
One hierarchical clustering involved only the rows containing local traits, while the other contained only 
rows corresponding to nonlocal traits.
We calculated the Euclidean distances between all row pairs (in the respective 
submatrices) before clustering. Hierarchical clustering was performed using the iterative complete 
linkage method. Initially, each row is assigned to its own cluster. The iterative 
algorithm then combines the two most similar clusters at each step. Distances between 
clusters were computed with the Lance-Williams dissimilarity update formula for complete 
linkage [@lance1967general]. Lastly, we arranged the rows according to optimal leaf ordering [@hahsler2008getting;@dendextend]. 
We then combined the
local and nonlocal traits into a single heatmap for each hotspot. 
To achieve coloring contrasts in the heatmaps, we
set the color spectrum limits at 0 and 5. 
We annotated rows with six pieces of information: 1. mediation LOD differences, 2. mediation 
LOD difference proportions, 3. univariate QTL peak positions, 4. estimated allele effects 
vector at the univariate peak position, 5. univariate LOD score, and 6. whether the row 
corresponds to a local trait. For clarity of presentation, we also indicated those columns that  
correspond to a local traits. We didn't examine the possibility that a local trait might 
mediate the relationship between a QTL and another local trait.

>>>>

In analyzing the data from @keller2018genetic, we first established criteria for identifying local and 
nonlocal transcripts at each of the five hotspots. We defined a "local trait" to be a transcript whose corresponding gene lay within the hotspot. We focused on those local 
traits that strongly associated with QTL within the hotspot. To limit the scope of our analyses, 
we limited local traits to those with univariate LOD greater than 15.

We defined a "nonlocal trait" to be one with univariate evidence of association (LOD at least 7.18, the genome-wide threshold)
with a QTL in the hotspot [@keller2018genetic]. Additionally, we required that the nonlocal trait's gene not lie on the hotspot's chromosome. We ignored transcripts for genes on the hotspot chromosome 
but outside the hotspot interval.

For each hotspot, we performed both pairwise pleiotropy tests and linear regression-based 
mediation analyses. For every pair of traits (local or nonlocal) at a hotspot, we calculated a pleiotropy test 
statistic[@boehm2019testing]. Due to the computational burden, we didn't determine bootstrap p-values for these 
tests. For every local trait - nonlocal trait pair, we also performed a linear 
regression-based mediation analysis [@chick2016defining,@keller2018genetic]. We describe both procedures below.

@@

For $n$ mice and two traits, $\Sigma$ is a $2n$ by $2n$ covariance matrix, while $V_g$ and $V_e$ are $2$ by $2$ covariance matrices (\@ref(eq:Sigmahat)). Estimates $\hat V_g$ and $\hat V_e$ are obtained via an expectation-maximization algorithm [@zhou2014efficient;@gemma2].

From the two-dimensional scan results, we calculated pleiotropy likelihood ratio test (LOD) statistics for each
pair of traits [@jiang1995multiple;@boehm2019testing;@boehm2019thesis]. Because of computing limitations, we didn't obtain bootstrap p-values for the pleiotropy tests. Instead,
we worked directly with the test statistic values.

@@

In all of our mediation analyses, the nonlocal traits are the targets (Y), while the local
transcripts serve as candidate mediators (M) of the QTL-target relationship. 
We only considered as potential mediators those local traits with strong univariate LODs.
In Equations~\ref{eq:model1} through \ref{eq:model4}, $X$ is a matrix of founder allele dosages,
$B$ is a matrix
of founder allele effects, $W$ is a matrix of covariates, $C$ is a matrix of covariate
effects, $M$ is a vector of values for a single candidate mediator (a local expression trait, in our case), $\beta$ is the effect
of the mediator on the complex trait (after adjusting for other covariates in the model), $1$ is a vector with all entries set to 1, $\alpha$
is a parameter that corresponds to the mean trait value (after adjusting for other covariates in the model), and $E$ is a vector of normally distributed random errors.

@@

We created heatmaps to examine patterns among pleiotropy test results in each hotspot. Our analysis used both 
local and nonlocal traits.
We represented the pleiotropy test statistics as a symmetric matrix, where each row was one expression
trait and each column was an expression trait. 
Excepting the cells on the diagonal, each cell, then, contained the pleiotropy test statistic
from the analysis involving the row's expression trait and the column's expression trait.
To create a heatmap, we first performed two hierarchical clusterings of rows. 
One hierarchical clustering involved only the rows for local traits, while the other contained only 
rows corresponding to nonlocal traits.
We calculated the Euclidean distances between all row pairs (in the respective 
submatrices) before clustering. Hierarchical clustering was performed using the iterative complete 
linkage method. Initially, each row is assigned to its own cluster. The iterative 
algorithm then combines the two most similar clusters at each step. Distances between 
clusters were computed with the Lance-Williams dissimilarity update formula for complete 
linkage [@lance1967general]. Lastly, we arranged the rows according to optimal leaf ordering [@hahsler2008getting;@dendextend]. 
We then combined the
local and nonlocal traits into a single heatmap for each hotspot. 
To achieve coloring contrasts in the heatmaps, we
set the color spectrum limits at 0 and 5. 
We annotated rows with six pieces of information: 1. mediation LOD differences, 2. mediation 
LOD difference proportions, 3. univariate QTL peak positions, 4. estimated allele effects 
vector at the univariate peak position, 5. univariate LOD score, and 6. whether the row 
corresponds to a local trait. For clarity of presentation, we also indicated those columns that  
correspond to a local traits. We didn't examine the possibility that a local trait might 
mediate the relationship between a QTL and another local trait.
